\textbf{The Illusion Game: A Novel Experimental Paradigm Provides Evidence for a General Factor of Visual Illusion Sensitivity and Personality Correlates}

### 2023-01-02 submission/R1/manuscript.Rmd
homogenize comments

abstract: |
  Visual illusions are a gateway to understand how we construct our experience of reality. Unfortunately, important questions remain open, such as the hypothesis of a common factor underlying the sensitivity to different types of illusions, as well as of personality correlates of illusion sensitivity. In this study, we used a novel parametric framework for visual illusions to generate 10 different classic illusions (Delboeuf, Ebbinghaus, Rod and Frame, Vertical-Horizontal, Zöllner, White, Müller-Lyer, Ponzo, Poggendorff, Contrast) varying in strength, embedded in a perceptual discrimination task. We tested the objective effect of the illusions on errors and response times, and extracted participant-level performance scores (n=250) for each illusion. Our results provide evidence in favour of a general factor underlying the sensitivity to different illusions (labelled Factor *i*). Moreover, we report a positive link between illusion sensitivity and personality traits such as Agreeableness, Honesty-Humility, and negative relationships with Psychoticism, Antagonism, Disinhibition, and Negative Affect.
keywords          : "visual illusions, illusion game, Pyllusion, personality, general factor"
wordcount         : "3826"
bibliography      : references.bib
floatsintext      : yes
linenumbers       : yes
draft             : no
mask              : no
figurelist        : yes
tablelist         : no
footnotelist      : no
classoption       : "man"
output            : papaja::apa6_pdf
csl: utils/scientific-reports.csl
header-includes:
  - \usepackage[labelfont=bf, font={scriptsize, color=gray}]{caption}
editor_options:
  chunk_output_type: console
---

@@

<!-- Bayes -->
Although the nature of the processes underlying illusion perception - whether related to low-level features of the visual processing system [@cretenoud2019; @gori2016] or to top-down influences [@teufel2018; @caporuscio2022] - remains debated, a growing body of literature proposes to conceptualize illusions under the Bayesian brain hypothesis [@friston2010free]. In this context, illusions are conceptualized as anomalous beliefs that result from giving ample weight to prior knowledge to minimize prediction error in the face of ambiguous incoming sensory evidence. The predictive coding account further provides an explanation regarding the observations from clinical populations. Certain dispositional traits or characteristics (e.g., psychoticism) are seen as driven by alterations in the system's metacognitive components [@adams2013computational], resulting in an underweighting of priors during perceptual inferences, and manifesting as a decreased sensitivity to illusions [@koethe2009binocular].

@@

Despite strong theoretical foundations and hypotheses, the empirical evidence remains scarce, clouded by methodological hurdles. For instance, one key challenge can be found in the difficulty of adapting visual illusions to an experimental setting, which typically requires the controlled modulation of the specific variables of interest. Instead, existing studies typically use only one or a small subset of illusion types, with few contrasting conditions, restricting the findings' generalizability [@cretenoud2020illusions; @cretenoud2019; @bressan2021most]. Moreover, conventional paradigms often focus on the participants' subjective experience, by asking them the extent to which they perceive two identical targets as different [@lanyi2022can], having them estimate the targets' physical properties [@coren1976empirical], or through the method of adjustment, which involves having them adjust the targets to perceptually match a reference stimulus [@mylniec2016; @grzeczkowski2018; @cretenoud2020individual; @cretenoud2021visual]. This reliance on meta-cognitive judgements about one's subjective experience likely distorts the measurand, limiting the ability to reliably obtain more direct and objective measures of illusion sensitivity [@skottun2014subjective]. 
While some recent efforts have some made to implement more empirically rigorous paradigms [@cretenoud2021individual], most of the applied manipulations only focus on varying the physical dimensions of the illusion's target features without modulating its contextual elements, hence limiting the variability in the illusory effects of the stimuli presented. Furthermore, such prior studies have typically generated stimuli whose targets’ physical attributes vary over a relatively narrow range, thus further constraining the reliability of their findings. As such, it is possible that the recent evidence reported against a common factor of illusions could be due to the low stimulus variance instead of a true reflection of a lack of common mechanism.

@@

```{r message=FALSE, warning=FALSE, fig.cap="The parametric framework for visual illusions (Makowski et al., 2021) applied to the Müller-Lyer illusion (above). Below are examples of stimuli showcasing the manipulation of two parameters, task difficulty and illusion strength.", out.width="100%"}
knitr::include_graphics("../figures/Figure1.png")
```

@@

```{r message=FALSE, warning=FALSE, out.width="100%", fig.cap="Ten different illusions were used as stimuli in a perceptual task, where participants had to answer as fast as possibly, without making errors, according to specific instructions. For each illusion type, two parameters were experimentally manipulated, 1) the task difficulty (e.g., how large was the difference between the bigger and the smaller red circles in the Delboeuf illusion), and 2) the illusion strength (e.g., the size of the black circles in the Ebbinghaus illusion)."}
knitr::include_graphics("../figures/Figure2.png")
```

@@

Participants were first given a brief demographic survey, which collected information regarding their age, gender, country of birth, ethnicity and highest attained education level. This was followed by a practice series of illusions, after which the first series of 10 illusion blocks was presented in a randomized order, with a further randomization of the stimuli order within each block. Following this first series of blocks, two personality questionnaires were administered, the *IPIP6* [24 items, @sibley2011] - measuring 6 "normal" personality traits (Extraversion, Openness, Conscientiousness, Agreeableness, Neuroticism and Honesty-Humility), and the *PID-5* [25 items, @hopwood2012] - measuring 5 "pathological" personality traits (Disinhibition, Antagonism, Detachment, Negative Affect and Psychoticism). Next, the second series of 10 illusion blocks was presented (with new randomized orders of blocks and trials). In total, each participant underwent 1340 trials of which they had to respond "as fast as possible without making errors" (i.e., an explicit double constraint to mitigate the inter-individual variability in the speed-accuracy trade off) by pressing the correct arrow key  (left/right, or up/down depending on the illusion type). For instance, in the Müller-Lyer block, participants had to answer which one of the upper or bottom target line was the longest. All trials were required to be completed within a single-session (total experiment duration: ~55 minutes). The task was implemented using *jsPsych* [@de2015jspsych] and was hosted on *Pavlovia* [(**https://pavlovia.org/**)](https://pavlovia.org/){.uri}. The set of instructions for each illusion type is available in the experiment code.

@@

We excluded 6 participants upon inspection of the average error rate (when close to 50%, suggesting random answers), and reaction time distribution (when implausibly fast relative to the typical RT distribution). For the remaining participants, we discarded blocks with more than 50% of errors (2.16% of trials), possibly indicating that instructions were misunderstood (e.g., participants focused on the shorter line instead of the longer one), and 0.76% trials with extreme response times (\< 125 ms or \> 4 SD above mean RT). Additionally, due to a technical issue, no personality data was recorded for the first eight participants.

@@

Finally, for each of the individual illusion sensitivity scores (10 illusion-specific factors and the general Factor *i*), we tested the effect of contextual variables (screen size, screen refresh rate), demographic variables (sex, education, age), and personality traits. It should be noted that the screen size is a proxy of the true screen size, measured using the number of pixels.

@@

```{r message=FALSE, warning=FALSE, out.width="100%", fig.cap="Top: the effect of illusion strength and task difficulty on the error rate and reaction time (RT) for each individual illusion. The solid line represents the General Additive Model (GAM), and the dashed line corresponds to its approximation via linear models. Descriptive data is shown with stacked dots (for which errors start from the top) and distributions for RTs. Negative values for illusion strength correspond to congruent (i.e., facilitating) illusion effects. Task difficulty (the objective difference between the targets of perceptual decision) levels are shown as colors, with lower values corresponding to harder trials. The results for each illusion are surrounded by 4 extreme examples of stimuli, corresponding to the hardest difficulty (on top) and the strongest illusion (on the right for incongruent illusions). Bottom: We extracted the effect slope of the illusion strength and its interaction with task difficulty for each participant. We fitted a Structural Equation Model (SEM) suggesting that these manifest variables group to first-level illusion-specific latent factors, which then load on a general factor of illusion sensitivity (Factor \\textit{i})."}
knitr::include_graphics("../figures/Figure3.png")
```

@@

```{r message=FALSE, warning=FALSE, fig.cap="The upper plots show the illusion sensitivity scores as a function of sex and age (solid lines indicate significant relationships). Bottom plots show the correlation between the general factor (Factor \\textit{i}) of illusion sensitivity (on the x-axes) and personality traits.", out.width="100%"}
knitr::include_graphics("../figures/Figure4.png")
```

>>>>

abstract: |
  Visual illusions are a gateway to understand how we construct our experience of reality. Unfortunately, important questions remain open, such as the hypothesis of a common factor underlying the sensitivity to different types of illusions, as well as of personality correlates of illusion sensitivity. In this study, we used a novel parametric framework for visual illusions to generate 10 different classic illusions (Delboeuf, Ebbinghaus, Rod and Frame, Vertical-Horizontal, Zöllner, White, Müller-Lyer, Ponzo, Poggendorff, Contrast) varying in strength, embedded in a perceptual discrimination task. We tested the objective effect of the illusions on errors and response times, and extracted participant-level performance scores (n=250) for each illusion. Our results provide evidence in favour of a general factor underlying the sensitivity to different illusions (labelled Factor *i*). Moreover, we report a positive link between illusion sensitivity and personality traits such as Agreeableness, Honesty-Humility, and negative relationships with Psychoticism, Antagonism, Disinhibition, and Negative Affect.
keywords          : "visual illusions, illusion game, Pyllusion, personality, general factor"
wordcount         : "4084"
bibliography      : references.bib
floatsintext      : yes
linenumbers       : yes
draft             : no
mask              : no
figurelist        : yes
tablelist         : no
footnotelist      : no
classoption       : "man"
output            : papaja::apa6_pdf
csl: utils/scientific-reports.csl
header-includes:
  - \usepackage[labelfont=bf, font={scriptsize, color=gray}]{caption}
editor_options:
  chunk_output_type: console
---

@@

<!-- Bayes -->
Although the nature of the processes underlying illusion perception - whether related to low-level features of the visual processing system [@cretenoud2019; @gori2016] or to top-down influences [@teufel2018; @caporuscio2022] - remains debated, a growing body of literature proposes to conceptualize illusions under the Bayesian brain hypothesis [@friston2010free]. In this context,illusions are conceptualized as **non-veridical perceptual experiences that result from giving ample weight to prior knowledge to minimize prediction error in the face of biasing sensory evidence.** The predictive coding account further provides an explanation regarding the observations from clinical populations. Certain dispositional traits or characteristics (e.g., psychoticism) are seen as driven by alterations in the system's metacognitive components [@adams2013computational], resulting in an underweighting of priors during perceptual inferences, and manifesting as a decreased sensitivity to illusions [@koethe2009binocular].

@@

Despite strong theoretical foundations and hypotheses, the empirical evidence remains scarce, clouded by methodological hurdles. For instance, one key challenge can be found in the difficulty of adapting visual illusions to an experimental setting, which typically requires the controlled modulation of the specific variables of interest. Instead, existing studies typically use only one or a small subset of illusion types, with few contrasting conditions, restricting the findings' generalizability [@cretenoud2020illusions; @cretenoud2019; @bressan2021most]. Moreover, conventional paradigms often focus on the participants' subjective experience, by asking them the extent to which they perceive two identical targets as different [@lanyi2022can], having them estimate the targets' physical properties [@coren1976empirical], **or through the method of adjustment, which involves having them adjust the targets to perceptually match a reference stimulus [@mylniec2016; @grzeczkowski2018; @cretenoud2020individual; @cretenoud2021visual].** This reliance on meta-cognitive judgements about one's subjective experience likely distorts the measurand, limiting the ability to reliably obtain more direct and objective measures of illusion sensitivity [@skottun2014subjective]. 
**While some recent efforts have some made to implement more empirically rigorous paradigms [@cretenoud2021individual], most of the applied manipulations only focus on varying the physical dimensions of the illusion's target features without modulating its contextual elements, hence limiting the variability in the illusory effects of the stimuli presented. Furthermore, such prior studies have typically generated stimuli whose targets’ physical attributes vary over a relatively narrow range, thus further constraining the reliability of their findings. As such, it is possible that the recent evidence reported against a common factor of illusions could be due to the low stimulus variance instead of a true reflection of a lack of common mechanism.**

@@

```{r message=FALSE, warning=FALSE, fig.cap="The parametric framework for visual illusions (Makowski et al., 2021) applied to the Müller-Lyer illusion (above). Below are examples of stimuli showcasing the manipulation of two parameters, task difficulty and illusion strength.", out.width="100%"}
knitr::include_graphics("../../figures/Figure1.png")
```

@@

```{r message=FALSE, warning=FALSE, out.width="100%", fig.cap="Ten different illusions were used as stimuli in a perceptual task, where participants had to answer as fast as possibly, without making errors, according to specific instructions. For each illusion type, two parameters were experimentally manipulated, 1) the task difficulty (e.g., how large was the difference between the bigger and the smaller red circles in the Delboeuf illusion), and 2) the illusion strength (e.g., the size of the black circles in the Ebbinghaus illusion)."}
knitr::include_graphics("../../figures/Figure2.png")
```

@@

Participants were first given a brief demographic survey, which collected information regarding their **age, gender, country of birth, ethnicity and highest attained education level**. This was followed by a practice series of illusions, after which the first series of 10 illusion blocks was presented in a randomized order, with a further randomization of the stimuli order within each block. Following this first series of blocks, two personality questionnaires were administered, the *IPIP6* [24 items, @sibley2011] - measuring 6 "normal" personality traits (Extraversion, Openness, Conscientiousness, Agreeableness, Neuroticism and Honesty-Humility), and the *PID-5* [25 items, @hopwood2012] - measuring 5 "pathological" personality traits (Disinhibition, Antagonism, Detachment, Negative Affect and Psychoticism). Next, the second series of 10 illusion blocks was presented (with new randomized orders of blocks and trials). In total, each participant underwent 1340 trials of which they had to respond "as fast as possible without making errors" (i.e., an explicit double constraint to mitigate the inter-individual variability in the speed-accuracy trade off) by pressing the correct arrow key  (left/right, or up/down depending on the illusion type). For instance, in the Müller-Lyer block, participants had to answer which one of the upper or bottom target line was the longest. **All trials were required to be completed within a single-session (total experiment duration: ~55 minutes)**. The task was implemented using *jsPsych* [@de2015jspsych] and was **hosted on *Pavlovia* ** [(**https://pavlovia.org/**)](https://pavlovia.org/){.uri}. The set of instructions for each illusion type is available in the experiment code.

@@

We excluded 6 participants upon inspection of the average error rate (when close to 50%, suggesting random answers), and reaction time distribution (when implausibly fast relative to the average RT distribution). For the remaining participants, we discarded blocks with more than 50% of errors (2.16% of trials), possibly indicating that instructions were misunderstood (e.g., participants focused on the shorter line instead of the longer one), and 0.76% trials with extreme response times (< 125 ms or > 4 SD above mean RT). Additionally, due to a technical issue, no personality data was recorded for the first eight participants.

@@

Finally, for each of the individual illusion sensitivity scores (10 illusion-specific factors and the general Factor *i*), we tested the effect of contextual variables (screen size, screen refresh rate), demographic variables (sex, education, age), and personality traits. It should be noted that the measure of screen size used (measured using the number of pixels) is **only a proxy of the true physical screen size**.

@@

```{r message=FALSE, warning=FALSE, out.width="100%", fig.cap="Top: the effect of illusion strength and task difficulty on the error rate and reaction time (RT) for each individual illusion. The solid line represents the General Additive Model (GAM), and the dashed line corresponds to its approximation via linear models. Descriptive data is shown with stacked dots (for which errors start from the top) and distributions for RTs. Negative values for illusion strength correspond to congruent (i.e., facilitating) illusion effects. Task difficulty (the objective difference between the targets of perceptual decision) levels are shown as colors, with lower values corresponding to harder trials. The results for each illusion are surrounded by 4 extreme examples of stimuli, corresponding to the hardest difficulty (on top) and the strongest illusion (on the right for incongruent illusions). Bottom: We extracted the effect slope of the illusion strength and its interaction with task difficulty for each participant. We fitted a Structural Equation Model (SEM) suggesting that these manifest variables group to first-level illusion-specific latent factors, which then load on a general factor of illusion sensitivity (Factor \\textit{i})."}
knitr::include_graphics("../../figures/Figure3.png")
```

@@

```{r message=FALSE, warning=FALSE, fig.cap="The upper plots show the illusion sensitivity scores as a function of sex and age (solid lines indicate significant relationships). Bottom plots show the correlation between the general factor (Factor \\textit{i}) of illusion sensitivity (on the x-axes) and personality traits.", out.width="100%"}
knitr::include_graphics("../../figures/Figure4.png")
```
